Taking Dyads Seriously

### 2018-02-05 paper/2_model.tex
modeling + sim section edits, recompile

To account for the dependencies that are prevalent in dyadic data, we turn to the AME model. The AME approach can be used to conduct inference on cross-sectional and longitudinal networks with binary, ordinal, or continuous linkages. It is flexible and easy to use for analyzing the kind of relational data often found in social science. It accounts for nodal and dyadic dependence patterns, as well as higher-order dependencies such as homophily and stochastic equivalence.\footnote{\citet{minhas:etal:2016:arxiv} detail how this framework contrasts with alternative network-based approaches.} The AME model accounts for higher-order dependencies using an approach similar to that of the recommender systems underlying the recommendation algorithms of companies such as Amazon \citep{resnick:varian:1997}.

The AME model combines the social relations regression model (SRRM) to account for nodal and dyadic dependencies and the latent factor model (LFM) for third-order dependencies.  For details on the SRRM see \citet{li:loken:2002,hoff:2005,dorff:minhas:2017}.\footnote{An earlier version of the LFM  used in AME is presented as the general bilinear mixed effects (GBME) model in \citet{hoff:ward:2004}. The GBME model is more limited in the types of dependence patterns that it can capture due to the formulation of the matrix decomposition procedure underlying the LFM.} The AME model is specified as follows:

@@

The LFM contribution to the AME comes in the multiplicative term: $\mathbf{u}_{i}^{\top} \mathbf{D} \mathbf{v}_{j}=\sum_{k \in K} d_{k} u_{ik} v_{jk}$. $K$ denotes the dimensions of the latent space. This model posits a latent vector of characteristics $\mathbf{u_{i}}$ and $\mathbf{v_{j}}$ for each sender $i$ and receiver $j$. The similarity or dissimilarity of these vectors will then influence the likelihood of activity, and provides a representation of third-order interdependencies \citep{minhas:etal:2016:arxiv}. The LFM parameters are estimated by a process similar to computing the singular value decomposition (SVD) of the observed network. When taking the SVD we factorize our observed network into the product of three matrices: $\mathbf{U}, \mathbf{D}, \text{ and }, \mathbf{V}$. This provides us with a low-dimensional representation of our original network.\footnote{The dimensions of $\mathbf{U}$ and $\mathbf{V}$ are $n \times K$ and $\mathbf{D}$ is a $K \times K$ diagonal matrix.} Values in $\mathbf{U}$ provide a representation of how stochastically equivalent actors are as senders in a network or, for example, how similar actors are in terms of who they initiate conflict with. $\hat{\mathbf{u}}_{i} \approx \hat{\mathbf{u}}_{j}$ would indicate that actor $i$ and $j$ initiate battles with similar third actors. $\mathbf{V}$ provide a similar representation but from the perspective of how similar actors are as receivers. The values in $\mathbf{D}$, a diagonal matrix, represent levels of homophily in the network.\footnote{Unlike traditional SVD, in the latent factor model, the singular values are not restricted to be positive, thus allowing us to account for both positive and negative homophily.}

Note that this model easily generalizes to the case, common in international relations, where interactions are unidrected (for example the presence of conflict or a bilateral investment treaty). In this case, $\rho$ is constrained to be 1, sender and receiver effects are identical for each observation, as are $\mathbf{u_{i}}$ and $\mathbf{v_{j}}$. In the application section, we show the applicability of the AME approach to both directed and undirected dyadic data.

>>>>

To account for the dependencies that are prevalent in dyadic data, we turn to the AME model. The AME approach can be used to conduct inference on cross-sectional and longitudinal networks with binary, ordinal, or continuous linkages. It is flexible and easy to use for analyzing the kind of relational data often found in social science. It accounts for nodal and dyadic dependence patterns, as well as higher-order dependencies such as homophily and stochastic equivalence.\footnote{\citet{minhas:etal:2016:arxiv} detail how this framework contrasts with alternative network-based approaches.} The AME model combines the social relations regression model (SRRM) to account for nodal and dyadic dependencies and the latent factor model (LFM) for third-order dependencies.  For details on the SRRM see \citet{li:loken:2002,hoff:2005,dorff:minhas:2017}.\footnote{An earlier version of the LFM  used in AME is presented as the general bilinear mixed effects (GBME) model in \citet{hoff:ward:2004}. The GBME model is more limited in the types of dependence patterns that it can capture due to the formulation of the matrix decomposition procedure.} The AME model is specified as follows:

@@

The LFM contribution to the AME comes in the multiplicative term: $\mathbf{u}_{i}^{\top} \mathbf{D} \mathbf{v}_{j}=\sum_{k \in K} d_{k} u_{ik} v_{jk}$. $K$ denotes the dimensions of the latent space. The construction of the LFM here is actually quite similar to the recommender systems that companies like Amazon and Netflix have used to model customer behavior \citep{resnick:varian:1997,bennett:etal:2007}. This model posits a latent vector of characteristics $\mathbf{u_{i}}$ and $\mathbf{v_{j}}$ for each sender $i$ and receiver $j$. The similarity or dissimilarity of these vectors will then influence the likelihood of activity, and provides a representation of third-order interdependencies. The LFM parameters are estimated by a process similar to computing the singular value decomposition (SVD) of the observed network. When taking the SVD we factorize our observed network into the product of three matrices: $\mathbf{U}, \mathbf{D}, \text{ and }, \mathbf{V}$. This provides us with a low-dimensional representation of our original network.\footnote{The dimensions of $\mathbf{U}$ and $\mathbf{V}$ are $n \times K$ and $\mathbf{D}$ is a $K \times K$ diagonal matrix.} Values in $\mathbf{U}$ provide a representation of how stochastically equivalent actors are as senders in a network or, for example, how similar actors are in terms of who they initiate conflict with. $\hat{\mathbf{u}}_{i} \approx \hat{\mathbf{u}}_{j}$ would indicate that actor $i$ and $j$ initiate battles with similar third actors. $\mathbf{V}$ provide a similar representation but from the perspective of how similar actors are as receivers. The values in $\mathbf{D}$, a diagonal matrix, represent levels of homophily in the network.\footnote{Unlike traditional SVD, in the latent factor model, the singular values are not restricted to be positive, thus allowing us to account for both positive and negative homophily.}

Note that this model easily generalizes to the case, common in IR, where interactions are undirected (for example the presence of conflict or a bilateral investment treaty). In the case of the SRRM, $\rho$ is constrained to be one and instead of separate sender and receiver random effects a single actor random effect is utilized. For the LFM, an eigendecomposition scheme is used to capture higher order dependence patterns. In the application section, we show the applicability of the AME approach to both directed and undirected dyadic data.

### 2018-02-05 paper/3_sim.tex
modeling + sim section edits, recompile

We utilize a simulation study to highlight the utility of AME as an inferential tool for dyadic analysis. The goal of the simulation is to assess how well AME can provide unbiased and well-calibrated estimates of coefficient parameters in the presence of unobserved dependencies. Specifically, we are concerned with conducting inference on regression parameters of a linear model for a network in the case where there is an omitted variable. For instance, assume that the true data-generating process for a particular $Y$ is given by:

@@

where $Y= \{y_{i,j}\}\in \mathbb R^{n\times n}$ is an observed sociomatrix, $X = \{x_{i,j} \} \in \mathbb R^{n \times n}$ is a matrix of observed dyad-specific characteristics, and $W = \{ w_{i,j}\} \in \mathbb R^{n \times n}$ is a matrix of unobserved dyad-specific characteristics. $Y$ can be thought of as a dyadic dependent variable, $X$ and $W$ are both dyadic covariates that are a part of the data-generating process for $Y$, but $W$ is not observed. We compare inference for $\mu$ and $\beta$---the latter parameter would be of primarily theoretical concern for applied scholars---using three models :

\begin{itemize}
	\item the standard international relations approach assuming independent errors; 
	\item the AME approach outlined in the previous section with a unidimensional latent factor space ($K=1$);
	\item and an ``oracle'' regression model that assumes we have measured all sources of dependencies and thus includes both $x_{i,j}$ and $w_{i,j}$. 
\end{itemize}

@@

We compare the performance of the models first in terms of how well they estimate the true values of $\mu$ and $\beta$ in Figure~\ref{fig:ameBias} by depicting the average $\mu$ and $\beta$ estimates from the simulations for the three models. The panels in the left show the results for when the number of actors is set to 50 and those on the right for 100; and the top pair of panels represents the estimates for $\mu$ while the bottom pair do the same for $\beta$. In each case, we find that the estimates for $\mu$ and $\beta$ produced by the standard approach are notably off from their true values. On the other hand, the AME model performs just as well as the oracle case in estimating the true values.

@@

Next, we estimate the 95\% confidence interval for the three models in each of the simulations and estimate the proportion of times that the true value fell within those intervals. The results are summarized in Figure~\ref{fig:ameCalib}, and again we see that the AME approach performs as well as the oracle, while the standard approach performs poorly by comparison.

@@

In the case of the data generating process for $Y$, $W$ is set as an unobserved dyadic covariate that had a homophilous effect on $Y$. Homophilous because $W$ within this framework is simply an example of a dyadic attribute involving $i$ and $j$ that positively affects the degree to which they will interact with one another, i.e., $y_{ij}$. This type of unobserved dependency will be captured through the multiplicative effects portion of the model, $\mathbf{U}^{\top} \mathbf{D} \mathbf{V}$. To estimate how well the model performs in doing this we recover the multiplicative effects term for each simulation and calculate the correlation between it and the unobserved dependency, $W$.\footnote{Specifically, since both the multiplicative effects term and $W$ are continuous dyadic variables, we calculate the Pearson correlation coefficient.} We visualize the distribution of the correlations from each of the 1,000 simulations in Figure~\ref{fig:ameCorr} for the case where the number of actors is set to 100 (top pair of panels) and 50. Additionally, we calculate the median across the correlations and display the result using a vertical line. For both $n=50$ and $n=100$, we find that the multiplicative effects perform very well in capturing the unobserved dependency, which indicates that the structure provided by this framework is not simply capturing noise but can be used as a tool to estimate unobserved structure.

@@

What this simulation has shown is that beyond obtaining less biased and better-calibrated parameter estimates, a key benefit of the AME framework is that we can also estimate unobserved dependencies through the random effects structure of the model.

% - in the presence of deendencies, ame can be used to easily perform accurate inference on exogenous parameters. the simulation here shows that the typial glm estimator has issues with bias and calibration
% - apart from just allowing inference on xbeta, the random effects are not just capturing random noise but underlying structure that we can further explore. we illustrate some strategies for doing so in teh application section.

>>>>

We utilize a simulation study to highlight the utility of AME as an inferential tool for dyadic analysis. Most scholars working with dyadic data are primarily concerned with understanding the effect of a particular independent variable on a dyadic dependent variable. The goal of the simulation is to assess how well AME can provide unbiased and well-calibrated estimates of coefficient parameters in the presence of unobserved dependencies. Specifically, we are concerned with conducting inference on regression parameters of a linear model for a network in the case where there is an omitted variable. For instance, assume that the true data-generating process for a particular $Y$ is given by:

@@

where $Y= \{y_{i,j}\}\in \mathbb R^{n\times n}$ is an observed sociomatrix, $X = \{x_{i,j} \} \in \mathbb R^{n \times n}$ is a matrix of observed dyad-specific characteristics, and $W = \{ w_{i,j}\} \in \mathbb R^{n \times n}$ is a matrix of unobserved dyad-specific characteristics. $Y$ can be thought of as a dyadic dependent variable, $X$ and $W$ are both dyadic covariates that are a part of the data-generating process for $Y$, but $W$ is not observed. We compare inference for $\mu$ and $\beta$---the latter parameter would be of primarily theoretical concern for applied scholars---using three models:

\begin{itemize}
	\item the ``standard'' international relations approach assuming independent errors; 
	\item the AME approach outlined in the previous section with a unidimensional latent factor space ($K=1$);\footnote{Results with higher values of $K$ are similar.}
	\item and an ``oracle'' regression model that assumes we have measured all sources of dependencies and thus includes both $x_{i,j}$ and $w_{i,j}$. 
\end{itemize}

@@

We compare the performance of the models first in terms of how well they estimate the true values of $\mu$ and $\beta$ in Figure~\ref{fig:ameBias} by depicting the average $\mu$ and $\beta$ estimates from the simulations for the three models. The panels in the left show the results for when the number of actors is set to 50 and those on the right for 100; and the top pair of panels represents the estimates for $\mu$ while the bottom pair do the same for $\beta$. In each case, we find that the estimates for $\mu$ and $\beta$ produced by the standard approach are notably off from their true values. On the other hand, the AME model performs just as well as the oracle case in estimating the true parameter values.

@@

Next, we estimate the 95\% confidence interval for the three models in each of the simulations and estimate the proportion of times that the true value fell within those intervals. The results are summarized in Figure~\ref{fig:ameCalib}, and again we see that the AME approach performs as well as the oracle, while the standard approach performs poorly by comparison. The implication of the results presented in Figures~\ref{fig:ameBias} and ~\ref{fig:ameCalib} is that standard approaches can often fail at estimating parameter values and conducting inferential tasks in the presence of unobserved dependencies. The AME approach by comparison can be used as a tool for scholars working with dyadic data to still estimate the true effects of their main variables of interest, while accounting for dependencies that do often emerge in dyadic data.

@@

Moreover, the AME approach allows scholars to better understand what parameters their model may be missing. In the case of the simulation here, $W$ is set as an unobserved dyadic covariate that had a homophilous effect on $Y$. Homophilous because $W$ within this framework is simply an example of a dyadic attribute involving $i$ and $j$ that positively affects the degree to which they will interact with one another, i.e., $y_{ij}$. This type of unobserved dependency will be captured through the multiplicative effects portion of the model, $\mathbf{U}^{\top} \mathbf{D} \mathbf{V}$. To estimate how well the model performs in doing this we recover the multiplicative effects term for each simulation and calculate the correlation between it and the unobserved dependency, $W$.\footnote{Specifically, since both the multiplicative effects term and $W$ are continuous dyadic variables, we calculate the Pearson correlation coefficient.} We visualize the distribution of the correlations from each of the 1,000 simulations in Figure~\ref{fig:ameCorr} for the case where the number of actors is set to 100 (top pair of panels) and 50. Additionally, we calculate the median across the correlations and display the result using a vertical line. For both $n=50$ and $n=100$, we find that the multiplicative effects perform very well in capturing the unobserved dependency, which indicates that the structure provided by this framework is not simply capturing noise but can be used as a tool to estimate unobserved structure.

@@

What this simulation has shown is that beyond obtaining less biased and better-calibrated parameter estimates, a key benefit of the AME framework is that we can also estimate unobserved dependencies through the random effects structure of the model. Scholars can use this framework in an iterative fashion where once they have estimated a model, they can empirically study the structure of the random effects portion of the model to assess whether there are unobserved covariates that they want to account for in their models. At the very least, what this simulation section shows is that the AME model can help in estimating the true effect and conducting inference on the independent variables that scholars have operationalized to test their theoretical propositions.

### 2018-02-05 paper/4_app.tex
modeling + sim section edits, recompile

We choose five recent studies from the broad field of international relations and international political economy that utilize relational data \citep{reiter:stam:2003, mcdonald:2004,  rose:2004, weeks:2012, gibler:2017}. These articles demonstrate the broad range of application for the AME framework, which include testing the effect of both dyadic and monadic variables on varied conflict behaviors and trade. It is worth noting, however, that the AME is applicable only where a full set of dyads for an outcome of interest is observable  -- for example it is unsuitable for studying onsets, where dyad-years representing conflict continuations are removed from the sample. The AME is not applicable to analyses of case-level data, for example, studies that examine the decision to go to war by particular states (e.g., the USA).

>>>>

We choose five recent studies from the broad field of international relations and international political economy that utilize relational data \citep{reiter:stam:2003, mcdonald:2004,  rose:2004, weeks:2012, gibler:2017}. These articles demonstrate the broad range of application for the AME framework, which include testing the effect of both dyadic and monadic variables on varied conflict behaviors and trade. It is worth noting, however, that the AME is applicable only where a full set of dyads for an outcome of interest is observable---for example it is unsuitable for studying onsets, where dyad-years representing conflict continuations are removed from the sample. The AME is not applicable to analyses of case-level data, for example, studies that examine the decision to go to war by particular states (e.g., the USA).
